---
title: "Topological Data Analysis in R"
author: "Jeremy Williams"
date: "June 4th, 2018"
output:
  html_document:
    fig_height: 4
    fig_width: 6
    toc: yes
    toc_depth: 2
  pdf_document:
    fig_height: 4
    fig_width: 6
    toc: yes
    toc_depth: 2
subtitle: Mathematics for Big Data (R Coding Report)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.align = "center")
```

\newpage

The *TDA* package in R Environment
========================================================

TDA stands for Topological Data Analysis. In particular, it provides functions for the statistical analysis of persistent homology and for density clustering. It includes the implementations of functions that, given some data, provide topological information about underlying space such as the distance function, the distance to measure, the kNN density estimator (knn), the kernel density estimator (kde), the kernel distance function and the density clustering using knn and kde.

One of the tools from algebraic topology used for TDA is called as Persistent homology. This tool allows us to detect large or small features in the data in the form of holes.

TDA package requires the following 4 packages as a prerequisite:

 1. **Parallel**: This package handles running much larger chunks of computations in parallel. 
 2. **FNN**: Fast Nearest Neighbour Search Algorithms and Applications.
 3. **igraph**: Provides functions for generating random and regular graphs, graph visualization etc.
 4. **scales**: Provides methods for automatically determining breaks and labels for axes and legends.
 
\newpage

R Environment Set Up
========================================================
Before installing **TDA** package, **parallel**, **FNN**, **igraph** and **scales** need to be installed. 

## Installing Packages (*parallel*, *FNN*, *igraph* and *scales*)
```{r}
if(!require(package = "parallel")){
  install.packages(pkgs = "parallel")
}
if(!require(package = "FNN")){
  install.packages(pkgs = "FNN")
}
if(!require(package = "igraph")){
  install.packages(pkgs = "igraph")
}
if(!require(package = "scales")){
  install.packages(pkgs = "scales")
}
```

## Installing *TDA* Package
```{r}
if(!require(package = "TDA")){
  install.packages(pkgs = "TDA")
}
```

## Other Packages (*rgl*, *gdata*, *readxl*, *formattable*, *ggplots*, *gplots*, *psych* and *pander*)
```{r}
if(!require(package = "rgl")){
  install.packages(pkgs = "rgl")
}
if(!require(package = "gdata")){
  install.packages(pkgs = "gdata")
}
if(!require(package = "readxl")){
  install.packages(pkgs = "readxl")
}
if(!require(package = "formattable")){
  install.packages(pkgs = "formattable")
}
if(!require(package = "ggplot2")){
  install.packages(pkgs = "ggplot2")
}
if(!require(package = "gplots")){
  install.packages(pkgs = "gplots")
}
if(!require(package = "psych")){
  install.packages(pkgs = "psych")
}
if(!require(package = "pander")){
  install.packages(pkgs = "pander")
}
```

## Loading Packages
```{r}
library(scales)
library(FNN)
library(igraph)
library(TDA)
library(rgl)
library(gdata)
library(readxl)
library(formattable)
library(ggplot2)
library(gplots)
library(psych)
library(pander)
```

\newpage

Breast Tissue Data Set
=======================================================
The Breat Tissue Data Set is freely available on the UCI Machine Learning Repository Website.

The Dataset can be downloaded here [link](http://archive.ics.uci.edu/ml/datasets/Breast+Tissue)

The Dataset has electrical impedance measurements of freshly excised breast tissue.

The Dataset is a multivariate dataset having 10 attributes (9 numeric types, 1 class type) and 106 instances.
	
* Dataset Attribute Information:
    + **Class**: Different types of breast cancer observed/studied 
    + **I0**: Impedivity at zero frequency
    + **PA500**: Phase angle at 500KHz
    + **HFS**: High frequency slope of phase angle
    + **DA**: Impedance distance between spectral ends
    + **Area**: Area under spectrum
    + **A/DA**: Area normalized by DA
    + **MAX IP**: Maximum of the spectrum
    + **DR**: Distance between I0 and the real part of the maximum frequency point
    + **P**: Length of the spectral curve

## Dataset Purpose

The Dataset can be used for predicting the classification of either the original 6 classes or of 4 classes by merging together the fibro-adenoma, mastopathy and glandular classes.

In this Report, the Dataset will be used to analyze two (2) attributes, **DA** and **DR**, using **TDA version 1.6.2** in **R Envirnoment**. 

Details for TDA version 1.6.2 can be viewed here [link](https://www.rdocumentation.org/packages/TDA/versions/1.6.2).

## Describe and Plot the Dataset (required packages are *readxl*, *psych* and *pander*)
```{r,fig.align = "center"}
data = read_excel("BreastData.xls")
attach(data)
ddata <-describe(data)

panderOptions("table.split.table", Inf)  ## don't split tables
pander(ddata)

Bdata<-cbind(data[,3],data[,4],data[,5],data[,6],data[,7],data[,8],data[,9])
plot(Bdata,main = "Breast Tissue Data", pch = 21, bg = rainbow(7))
```

\newpage

Basic Functions from TDA for data generation.
========================================================
## Circle
```{r,fig.align = "center"}
circle = circleUnif(n = 100, r = 2)       # r is the radius
plot3d(circle,col = rainbow(7))
plot(circle, pch = 21, bg = rainbow(7))
```

## Sphere        (required package is *rgl*)
```{r,fig.align = "center"}
 
sphere = sphereUnif(n = 1000, d = 2, r = 2) # r is the radius, 
                                            # d is dimension of real space minus 1.
plot3d(sphere, col = rainbow(7))
plot(sphere,pch = 21, bg = rainbow(7))
```

## Torus       (required package is *rgl*)
```{r,fig.align = "center"}
torus = torusUnif(n = 10000, a= 1.5, c = 3)
plot3d(torus,col = rainbow(7))
plot(torus,pch = 21, bg = rainbow(7))
```

\newpage

DA & DR Tables with Model and Diagnostic Plots 
========================================================

```{r,fig.align = "center"}
DA <- data$DA
DR <- data$DR
datap <- data.frame(DA, DR)


panderOptions("table.split.table", Inf)  ## don't split tables
pander(summary(datap))
```


# Model (DA ~ DR) 
```{r,fig.align = "center"}
plot(data$DA, data$DR, 
     xlab="Impedance distance between spectral ends", 
     ylab="Distance between I0 and maximum frequency point",
     main="Normal (DA ~ DR)",pch = 21, bg = rainbow(2))
lines(lowess(data$DA,data$DR), col="blue") # lowess line (x,y)
grid()

plotLowess(data$DA ~ data$DR, data=data, 
           xlab="Distance between I0 and maximum frequency point",
           ylab="Impedance distance between spectral ends", 
           main="Lowess (DR ~ DA)",pch = 21, bg = rainbow(2))
```

\newpage

Area, DA & DR Tables with Model and Diagnostic Plots
========================================================

```{r,fig.align = "center"}
Area <-data$Area
DA <- data$DA
DR <- data$DR

datapm <- data.frame(Area, DA, DR)

panderOptions("table.split.table", Inf)  ## don't split tables
pander(summary(datapm))

```

# Prediction of Model (Area ~ DA + DR) 
```{r,fig.align = "center"}
# Fit a linear model and run a summary of its results.

Area <-data$Area 
DA <-data$DA
DR <-data$DR

mod1<-lm(Area ~ DA + DR, data = data)

panderOptions("table.split.table", Inf)  ## don't split tables
pander(summary(mod1))

# Predicted values 
Predict_data <- fitted(mod1)
plot(Predict_data, main="Predicted Values",pch = 21, bg = rainbow(2))
grid()

# Model coefficients
panderOptions("table.split.table", Inf)  ## don't split tables
pander(coefficients(mod1))


# CIs for model parameters
panderOptions("table.split.table", Inf)  ## don't split tables
pander(confint(mod1, level=0.95))


# Plot model residuals
plot(mod1, pch=21, which=1, bg = rainbow(2))
grid()

# Extract coefficients
mod1_coef <- round(coef(mod1), 3) 

# Display equation 
mtext(bquote(Area == .(mod1_coef[3])* DR +.(mod1_coef[2])* DA + .(mod1_coef[1])),
      adj=1, padj=-2) 
```

\newpage

Distance Functions and Density Estimators
========================================================
```{r,fig.align = "center"}
X = data.frame(DA, DR)
Xlim = c(-10000, 10000)
Ylim = c(-10000, 10000)
by = 27
Xseq = seq(from = Xlim[1], to = Xlim[2], by = by)
Yseq = seq(from = Ylim[1], to = Ylim[2], by = by)
Grid = expand.grid(Xseq, Yseq)        #Generating the grid.
distance = distFct(X = X, Grid = Grid)

par(mfrow = c(1,2))
plot(X, xlab = "DA", ylab = "DR", main = "DA vs DR",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(distance, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -30, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "Distance Function")
```

\newpage

Distance to Measure (DTM)
========================================================
```{r,fig.align = "center"}
smoothing_parameter <- 0.2
DTM <- dtm(X = X, Grid = Grid, m0 = smoothing_parameter)

par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(DTM, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -30, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "DTM")
```

\newpage

KNN Density Estimator
========================================================
```{r,fig.align = "center"}
k <- 60
kNN <- knnDE(X = X, Grid = Grid, k = k)

par(mfrow = c(1,2))
plot(X, xlab = "DA", ylab = "DR", main = "Sample X",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(kNN, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -35, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "kNN")
```

\newpage

Gaussian Kernel Density Estimator (KDE)
========================================================
```{r,fig.align = "center"}
h <- 0.1
KDE <- kde(X = X, Grid = Grid, h = h)

par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(kNN, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 50, shade = 0.5,
main = "KDE")
```

\newpage

Kernel Distance
========================================================
```{r,fig.align = "center"}
h <- 0.3
Kdist <- kernelDist(X = X, Grid = Grid, h = h)

par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(Kdist, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 50, shade = 0.5,
main = "Kernel Distance")
```

\newpage

Bootstrap Confidence Bands and Persistent Homology 
========================================================
```{r,fig.align = "center"}
band <- bootstrapBand(X = X, FUN = kde, Grid = Grid, B = 100,
parallel = FALSE, alpha = 0.1, h = h)

pander(band[["width"]])
```
```{r,fig.align = "center"}
Diag <- gridDiag(X = X, FUN = kde, lim = cbind(Xlim, Ylim), by = by,
sublevel = FALSE, library = "Dionysus", printProgress = FALSE, h = 0.3)
par(mfrow = c(1,3))
plot(X, main = "Sample X",pch = 21, bg = rainbow(2))

persp(x = Xseq, y = Yseq,
z = matrix(KDE, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 3, col = "red", border = NA, ltheta = 50, shade = 0.9,
main = "KDE")

plot(x = Diag[["diagram"]], band = 2 * band[["width"]], main = "KDE Diagram")
```

\newpage

Rips Diagram 
========================================================
```{r,fig.align = "center"}
X1 = X
X2 = X + sd(X$DA) + sd(X$DR)

pander(X2)

X_final = rbind(X1, X2)

pander(X_final)

par(mfrow = c(1,1))
plot(X_final, xlab = "", ylab = "",pch = 21, bg = rainbow(2))
```
```{r}
Diag <- ripsDiag(X = X_final, maxdimension = 1, maxscale = 5,
library = "GUDHI", printProgress = FALSE)

par(mfrow=c(1,2))
plot(X_final, xlab="", ylab="",pch = 21, bg = rainbow(2))
plot(Diag[["diagram"]])
```
```{r}
Diag1 <- ripsDiag(X = X1, maxdimension = 1, maxscale = 5)
Diag2 <- ripsDiag(X = X2, maxdimension = 1, maxscale = 5)

pander(bottleneck(Diag1 = Diag1[["diagram"]], Diag2 = Diag2[["diagram"]],
dimension = 1))

```
```{r}
pander(wasserstein(Diag1 = Diag1[["diagram"]], Diag2 = Diag2[["diagram"]],
p = 2, dimension = 1))
```


\newpage

Density Clustering 
========================================================

```{r,fig.align = "center"}
Y = data.frame(DA, DR)
Tree <- clusterTree(Y, k =6,density = "knn", printProgress = FALSE)
plot(Tree, type = "lambda", main = "lambda Tree (kNN)")
plot(Tree, type = "kappa", main = "kappa Tree (kNN)")
```
```{r}
TreeKDE = clusterTree(Y, k = 4, h = 0.3, density = "kde", printProgress = FALSE)
plot(TreeKDE, type = "lambda", main = "lambda TreeKDE (kNN)")
plot(TreeKDE, type = "kappa", main = "kappa TreeKDE (kNN)")
```


\newpage

Conclusion 
========================================================
Biological phenomena are intrinsically highly variable and stochastic in nature, and notions of biological similarity are less rigid. Consequently, analysis methods for biomedical data need to identify shape characteristics that are fairly robust to changes by rescaling of distances and therefore become more qualitative in nature. This has led us to use methods adapted from the mathematics area of topology, which studies precisely the characteristics of shapes that are not rigid. 

In this report, TDA in R Environment was performed on breast cancer data. We started with plotting the data and applying basic TDA functions on the data. The variables chosen for analysis is DA which is the impedance distance between spectral ends and DR which is distance between I0 and real part of the maximum frequency point. These measurements plotted in the (real, -imaginary) plane constitute the impedance spectrum from where the breast tissue features are computed. Our aim was to identify the different types of breast cancers based on the variability of impedivity in them. Distance function and density estimators were applied on the data and a linear relation between DA and DR could be seen. Similar graph was seen for the distance to measure graph. KNN density estimator, kernel and Gaussian density estimator showed a flat graph. Bootstrap confidence bands and persistent homology analysis showed a band width of 90%. The density clustering analysis was performed and kappa and lambda trees were plotted. 

I believe that TDA, a group of methods for studying data from many different sources and of many different kinds, is particularly appropriate for the analysis of breast tissue data and can reveal interesting patterns. These methods begin the process of uncovering the topology or special organization of genomic data sets. TDA provides a viewpoint of these data which is combinatorial and therefore easy to grasp, and it has a degree of robustness to the sort of distortions that can occur in studying biomedical data.


\newpage

Reference 
========================================================

**Dua, D. and Karra Taniskidou, E.** (2017). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science. 

**Brittany T. Fasy, Jisu Kim, Fabrizio Lecci, Clement Maria, Vincent Rouvreau**, Statistical Tools for Topological Data Analysis, The Comprehensive R Archive Network (2018). https://cran.r-project.org/web/packages/TDA/TDA.pdf 

**Jossinet J** (1996) Variability of impedivity in normal and pathological breast tissue. Med. & Biol. Eng. & Comput, 34: 346-350.

**Silva JE, Marques de Sá JP, Jossinet J** (2000) Classification of Breast Tissue by Electrical Impedance Spectroscopy. Med & Bio Eng & Computing, 38:26-30.

**Nicolau, M., Levine, A. J. & Carlsson, G.** Topology based data analysis identifies a subgroup of breast cancers with a unique mutational profile and excellent survival. Proc Natl Acad Sci U S A 108, 7265-70 (2011).

**Madsen EVE, van Dalen J, van Gorp J, van Oort PMP, van Dalen T.** Frozen section analysis of sentinel lymph nodes in patients with breast cancer does not impair the probability to detect lymph node metastases. Virchows Arch. 2012;460:69-76. Published online 2011 Nov 25. doi: 10.1007/s00428-011-1171-1. PMCID:PMC3267028

**Czes Kosniowski**, A first course in algebraic topology. Cambridge University Press (1980).

**James R. Munkres**, Topology, Pentice Hall (2000).

**Allen Hatcher**, Algebraic Topology, (2001) https://www.math.cornell.edu/~hatcher/AT/AT.pdf

**Herbert Edelsbrunner**, A Short Course in Computational Geometry and Topology, Springer (2014)
