---
title: "Topological Data Analysis in R"
author: "Jeremy Williams"
date: "June 1st, 2018"
output:
  pdf_document:
    fig_height: 4
    fig_width: 6
    toc: yes
    toc_depth: 2
  html_document:
    fig_height: 4
    fig_width: 6
    toc: yes
    toc_depth: 2
subtitle: Mathematics for Big Data (R Coding Report)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

\newpage

The *TDA* package in R Environment
========================================================

TDA stands for Topological Data Analysis. In particular, it provides functions for the statistical analysis of persistent homology and for density clustering. It includes the implementations of functions that, given some data, provide topological information about underlying space such as the distance function, the distance to measure, the kNN density estimator (knn), the kernel density estimator (kde), the kernel distance function and the density clustering using knn and kde.

One of the tools from algebraic topology used for TDA is called as Persistent homology. This tool allows us to detect large or small features in the data in the form of holes.

TDA packages requires the following 4 packages as a prerequisite:

 1. **Parallel**: This package handles running much larger chunks of computations in parallel. 
 2. **FNN**: Fast Nearest Neighbour Search Algorithms and Applications.
 3. **igraph**: Provides functions for generating random and regular graphs, graph visualization etc.
 4. **scales**: Provides methods for automatically determining breaks and labels for axes and legends.
 
\newpage

R Environment Set Up
========================================================
Before installing **TDA** package, **parallel**, **FNN**, **igraph** and **scales** needs to be installed. 

## Installing Packages (*parallel*, *FNN*, *igraph* and *scales*)
```{r}
if(!require(package = "parallel")){
  install.packages(pkgs = "parallel")
}
if(!require(package = "FNN")){
  install.packages(pkgs = "FNN")
}
if(!require(package = "igraph")){
  install.packages(pkgs = "igraph")
}
if(!require(package = "scales")){
  install.packages(pkgs = "scales")
}
```

## Installing *TDA* Package
```{r}
if(!require(package = "TDA")){
  install.packages(pkgs = "TDA")
}
```

## Other Packages (*rgl*, *gdata*, *readxl*, *formattable*, *ggplots* and *gplots*)
```{r}
if(!require(package = "rgl")){
  install.packages(pkgs = "rgl")
}
if(!require(package = "gdata")){
  install.packages(pkgs = "gdata")
}
if(!require(package = "readxl")){
  install.packages(pkgs = "readxl")
}
if(!require(package = "formattable")){
  install.packages(pkgs = "formattable")
}
if(!require(package = "ggplot2")){
  install.packages(pkgs = "ggplot2")
}
if(!require(package = "gplots")){
  install.packages(pkgs = "gplots")
}
if(!require(package = "psych")){
  install.packages(pkgs = "psych")
}
```

## Loading Packages
```{r}
library(scales)
library(FNN)
library(igraph)
library(TDA)
library(rgl)
library(gdata)
library(readxl)
library(formattable)
library(ggplot2)
library(gplots)
library(psych)
```

\newpage

Breast Tissue Data Set
=======================================================
The Breat Tissue Data Set is freely available on the UCI Machine Learning Repository Website.

The Dataset can be downloaded here [link](http://archive.ics.uci.edu/ml/datasets/Breast+Tissue)

The Dataset has electrical impedance measurements of freshly excised breast tissue.

The Dataset is a multivariate dataset having 10 attributes (9 numeric types, 1 class type) and 106 instances.
	
* Dataset Attribute Information:
    + **Class**: Different types of breast cancer observed/studied 
    + **I0**: Impedivity at zero frequency
    + **PA500**: Phase angle at 500KHz
    + **HFS**: High frequency slope of phase angle
    + **DA**: Impedance distance between spectral ends
    + **Area**: Area under spectrum
    + **A/DA**: Area normalized by DA
    + **MAX IP**: Maximum of the spectrum
    + **DR**: Distance between I0 and the real part of the maximum frequency point
    + **P**: Length of the spectral curve

## Dataset Purpose

The Dataset can be used for predicting the classification of either the original 6 classes or of 4 classes by merging together the fibro-adenoma, mastopathy and glandular classes.

In this Report, the Dataset will be used to analyze two (2) attributes, **DA** and **DR**, using **TDA version 1.6.2** in **R Envirnoment**. 

Details for TDA version 1.6.2 can be viewed here [link](https://www.rdocumentation.org/packages/TDA/versions/1.6.2).

## Describe the Dataset (required packages are *readxl* and *psych*)
```{r}
data = read_excel("BreastData.xls")
attach(data)
describe(data)
```

\newpage

Basic Functions from TDA for data generation.
========================================================
## Circle
```{r}
names(data)
circle = circleUnif(n = 100, r = 2)       # r is the radius
plot(circle)
```

## Sphere        (required package is *rgl*)
```{r}
sphere = sphereUnif(n = 1000, d = 2, r = 2)       # r is the radius, d is dimension of real space minus 1. 
plot3d(sphere)
plot(sphere)
```

## Torus       (required package is *rgl*)
```{r}
torus = torusUnif(n = 10000, a= 1.5, c = 3)
plot3d(torus)
plot(torus)
```

\newpage

DA & DR Tables with Model and Diagnostic Plots 
========================================================

```{r}
DA <- data$DA
DR <- data$DR
datap <- data.frame(DA, DR)
summary(datap)
```


# Model (DA ~ DR) 
```{r}
plot(data$DA, data$DR, 
     xlab="Impedance distance between spectral ends", 
     ylab="Distance between I0 and maximum frequency point",
     main="Normal (DA ~ DR)")
lines(lowess(data$DA,data$DR), col="blue") # lowess line (x,y)
grid()

plotLowess(data$DA ~ data$DR, data=data, 
           xlab="Distance between I0 and maximum frequency point",
           ylab="Impedance distance between spectral ends", 
           main="Lowess (DR ~ DA)")
```

\newpage

Area, DA & DR Tables with Model and Diagnostic Plots
========================================================

```{r,message=FALSE, warning=FALSE}
Area <-data$Area
DA <- data$DA
DR <- data$DR

datapm <- data.frame(Area, DA, DR)
summary(datapm)
```

# Prediction of Model (Area ~ DA + DR) 
```{r}
# Fit a linear model and run a summary of its results.

Area <-data$Area 
DA <-data$DA
DR <-data$DR

mod1<-lm(Area ~ DA + DR, data = data)
summary(mod1)

# Predicted values 
Predict_data <- fitted(mod1)
plot(Predict_data, main="Predicted Values")
#lines(Predict_data, col="dark green")
grid()

# Model coefficients
coefficients(mod1)

# CIs for model parameters
confint(mod1, level=0.95)

# Plot model residuals
plot(mod1, pch=16, which=1)
grid()

# Extract coefficients
mod1_coef <- round(coef(mod1), 3) 

# Display equation 
mtext(bquote(Area == .(mod1_coef[3])* DR +.(mod1_coef[2])* DA + .(mod1_coef[1])),
      adj=1, padj=-2) 
```


\newpage

Distance Functions and Density Estimators
========================================================
```{r}
X = data.frame(DA, DR)
Xlim = c(-10000, 10000)
Ylim = c(-10000, 10000)
by = 27
Xseq = seq(from = Xlim[1], to = Xlim[2], by = by)
Yseq = seq(from = Ylim[1], to = Ylim[2], by = by)
Grid = expand.grid(Xseq, Yseq)        #Generating the grid.
distance = distFct(X = X, Grid = Grid)
par(mfrow = c(1,2))
plot(X, xlab = "DA", ylab = "DR", main = "DA vs DR")
persp(x = Xseq, y = Yseq,
z = matrix(distance, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -30, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "Distance Function")
```

\newpage

Distance to Measure (DTM)
========================================================
```{r}
smoothing_parameter <- 0.2
DTM <- dtm(X = X, Grid = Grid, m0 = smoothing_parameter)
par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X")
persp(x = Xseq, y = Yseq,
z = matrix(DTM, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -30, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "DTM")
```

\newpage

KNN Density Estimator
========================================================
```{r}
k <- 60
kNN <- knnDE(X = X, Grid = Grid, k = k)
par(mfrow = c(1,2))
plot(X, xlab = "DA", ylab = "DR", main = "Sample X")
persp(x = Xseq, y = Yseq,
z = matrix(kNN, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -35, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 30, shade = 0.5,
main = "kNN")
```

\newpage

Gaussian Kernel Density Estimator (KDE)
========================================================
```{r}
h <- 0.1
KDE <- kde(X = X, Grid = Grid, h = h)
par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X")
persp(x = Xseq, y = Yseq,
z = matrix(kNN, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 50, shade = 0.5,
main = "KDE")
```

\newpage

Kernel Distance
========================================================
```{r}
h <- 0.3
Kdist <- kernelDist(X = X, Grid = Grid, h = h)
par(mfrow = c(1,2))
plot(X, xlab = "", ylab = "", main = "Sample X")
persp(x = Xseq, y = Yseq,
z = matrix(Kdist, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 1.2, col = "red", border = NA, ltheta = 50, shade = 0.5,
main = "Kernel Distance")
```

\newpage

Bootstrap Confidence Bands and Persistent Homology 
========================================================
```{r}
band <- bootstrapBand(X = X, FUN = kde, Grid = Grid, B = 100,
parallel = FALSE, alpha = 0.1, h = h)
print(band[["width"]])
Diag <- gridDiag(X = X, FUN = kde, lim = cbind(Xlim, Ylim), by = by,
sublevel = FALSE, library = "Dionysus", printProgress = FALSE, h = 0.3)
par(mfrow = c(1,3))
plot(X, main = "Sample X")
persp(x = Xseq, y = Yseq,
z = matrix(KDE, nrow = length(Xseq), ncol = length(Yseq)),
xlab = "", ylab = "", zlab = "", theta = -20, phi = 35, scale = FALSE,
expand = 3, col = "red", border = NA, ltheta = 50, shade = 0.9,
main = "KDE")
plot(x = Diag[["diagram"]], band = 2 * band[["width"]], main = "KDE Diagram")
```

\newpage

Rips Diagram 
========================================================
```{r}
X1 = X
X2 = X + sd(X$DA) + sd(X$DR)
print(X2)
X_final = rbind(X1, X2)
print(X_final)
par(mfrow = c(1,1))
plot(X_final, xlab = "", ylab = "")
```
```{r}
Diag <- ripsDiag(X = X_final, maxdimension = 1, maxscale = 5,
library = "GUDHI", printProgress = FALSE)
par(mfrow=c(1,2))
plot(X_final, xlab="", ylab="")
plot(Diag[["diagram"]])
```
```{r}
Diag1 <- ripsDiag(X = X1, maxdimension = 1, maxscale = 5)
Diag2 <- ripsDiag(X = X2, maxdimension = 1, maxscale = 5)
print(bottleneck(Diag1 = Diag1[["diagram"]], Diag2 = Diag2[["diagram"]],
dimension = 1))

```
```{r}
print(wasserstein(Diag1 = Diag1[["diagram"]], Diag2 = Diag2[["diagram"]],
p = 2, dimension = 1))
```


\newpage

Density Clustering 
========================================================

```{r}
Y = data.frame(DA, DR)
Tree <- clusterTree(Y, k =6,density = "knn", printProgress = FALSE)
plot(Tree, type = "lambda", main = "lambda Tree (kNN)")
plot(Tree, type = "kappa", main = "kappa Tree (kNN)")
```
```{r}
TreeKDE = clusterTree(Y, k = 4, h = 0.3, density = "kde", printProgress = FALSE)
plot(TreeKDE, type = "lambda", main = "lambda TreeKDE (kNN)")
plot(TreeKDE, type = "kappa", main = "kappa TreeKDE (kNN)")
```

\newpage

Reference 
========================================================

**Dua, D. and Karra Taniskidou, E.** (2017). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science. 

**Brittany T. Fasy, Jisu Kim, Fabrizio Lecci, Clement Maria, Vincent Rouvreau**, Statistical Tools for Topological Data Analysis, The Comprehensive R Archive Network (2018). https://cran.r-project.org/web/packages/TDA/TDA.pdf 

**Jossinet J** (1996) Variability of impedivity in normal and pathological breast tissue. Med. & Biol. Eng. & Comput, 34: 346-350.

**Silva JE, Marques de Sá JP, Jossinet J** (2000) Classification of Breast Tissue by Electrical Impedance Spectroscopy. Med & Bio Eng & Computing, 38:26-30.


**Czes Kosniowski**, A first course in algebraic topology. Cambridge University Press (1980).

**James R. Munkres**, Topology, Pentice Hall (2000).

**Allen Hatcher**, Algebraic Topology, (2001) https://www.math.cornell.edu/~hatcher/AT/AT.pdf

**Herbert Edelsbrunner**, A Short Course in Computational Geometry and Topology, Springer (2014)
